---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# latenetwork: Inference on Local Average Treatment Effect Type Parameters under Network Interference of Unknown Form

<!-- badges: start -->
<!-- badges: end -->

The **latenetwork** package provides tools for causal inference 
under noncompliance with treatment assignment and 
network interference of unknown form.
The package enables to implement the instrumental variables (IV)
estimation for the local average treatment effect (LATE) type parameters 
via inverse probability weighting (IPW) using
the concept of instrumental exposure mapping (IEM)
and the framework of approximate neighborhood interference (ANI).
For more details,
see [Hoshino and Yanagi (2023) "Causal inference with noncompliance and unknown interference"](https://doi.org/10.48550/arXiv.2108.07455).

## Installation

Get the package from GitHub:

```{r, eval = F, echo = T}
# install.packages("devtools") # if necessary
devtools::install_github("tkhdyanagi/latenetwork", build_vignettes = TRUE)
```

## Vignettes

For more details, see the package vignettes with:

```{r, eval = FALSE}
library("latenetwork")

# Getting Started with the latenetwork Package
vignette("latenetwork")

# Review of Causal Inference with Noncompliance and Unknown Interference
vignette("review", package = "latenetwork")
```

## References

- Hoshino, T. and Yanagi, T., 2023. 
Causal inference with noncompliance and unknown interference. 
arXiv preprint arXiv:2108.07455.
[Link](https://doi.org/10.48550/arXiv.2108.07455)
